Selection on early survival does not explain germination rate clines in \emph{Mimulus cardinalis}

### 2022-07-10 ms/ms.tex
verified make
archive

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Selection on early survival does not explain germination rate clines in Mimulus cardinalis (Phrymaceae)},
  pdfauthor={Christopher D. Muir\^{}\{1,2,*\}, Courtney L. Van Den Elzen\^{}\{1,3\}, Amy L. Angert\^{}\{1,4\}},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{geometry}
\geometry{verbose,letterpaper,margin=2.45cm}

@@

\usepackage{setspace}
\doublespacing

@@

\title{Selection on early survival does not explain germination rate clines in \emph{Mimulus cardinalis} (Phrymaceae)}
\author{Christopher D. Muir\(^{1,2,*}\), Courtney L. Van Den Elzen\(^{1,3}\), Amy L. Angert\(^{1,4}\)}
\date{}

@@

Populations within species differ in ecologically important traits that often result from adaptation to different local environments (\protect\hyperlink{ref-turesson_genotypic_1922}{Turesson, 1922}; \protect\hyperlink{ref-clausen_experimental_1948}{Clausen et al., 1948}). However, it is still rare that we understand the traits and selective agents underlying local adaptation (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}). The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment (\protect\hyperlink{ref-kawecki_conceptual_2004}{Kawecki and Ebert, 2004}; \protect\hyperlink{ref-johnson_reciprocal_2021}{Johnson et al., 2021}), such that populations have higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower \emph{Mimulus guttatus} have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes (\protect\hyperlink{ref-lowry_widespread_2010}{Lowry and Willis, 2010}). The prevalence of local adaptation implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales (e.g. \protect\hyperlink{ref-wright_indirect_2013}{Wright et al., 2013}; \protect\hyperlink{ref-grant_40_2014}{Grant and Grant, 2014}; \protect\hyperlink{ref-richardson_microgeographic_2014}{Richardson et al., 2014}; \protect\hyperlink{ref-lescak_evolution_2015}{Lescak et al., 2015}; \protect\hyperlink{ref-barrett_linking_2019}{Barrett et al., 2019}; \protect\hyperlink{ref-divittorio_natural_2020}{DiVittorio et al., 2020}) likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent (\protect\hyperlink{ref-leimu_meta-analysis_2008}{Leimu and Fischer, 2008}; \protect\hyperlink{ref-hereford_quantitative_2009}{Hereford, 2009}; \protect\hyperlink{ref-brady_understanding_2019}{Brady et al., 2019}) and may be getting weaker because of anthropogenic climate change (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). For example, in the alpine plant \emph{Boechera strica}, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations (\protect\hyperlink{ref-anderson_climate_2020}{Anderson and Wadgymar, 2020}). Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

@@

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}), but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ between environments. If genetic variation in a trait \emph{causes} fitness to vary, then genotypic selection analyses (\protect\hyperlink{ref-rausher_measurement_1992}{Rausher, 1992}) should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of \emph{Arabidopsis thaliana} and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions (\protect\hyperlink{ref-postma_early_2016}{Postma and Ågren, 2016}). Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those that may have evolved nonadaptively.

@@

\emph{Mimulus cardinalis} (\(=\)\emph{Erythranthe cardinalis} \protect\hyperlink{ref-lowry_case_2019}{Lowry et al., 2019}) is a perennial forb native to the Western US (California and Oregon). It is self-compatible, predominantly outcrossing, and hummingbird pollinated. We used five focal populations from throughout the geographic range of \emph{M. cardinalis} (Table \ref{tab:focal_pops}; Fig. \ref{fig:range-map}). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits (\protect\hyperlink{ref-muir_grow_2017}{Muir and Angert, 2017}). Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for \(2-4\) weeks to dry, then stored at room temperature. To minimize maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4265}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1471}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1765}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
\caption{\label{tab:focal_pops} Source populations, including the name of the drainage where the seeds were collected, the latitude, longitude, and elevation in meters above sea level (mas).}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Source population
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Latitude
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Longitude
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Elevation (mas)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Source population
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Latitude
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Longitude
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Elevation (mas)
\end{minipage} \\
\midrule
\endhead
Sweetwater River & \(32.900\) & \(-116.585\) & \(1180\) \\
West Fork Mojave River & \(34.284\) & \(-117.378\) & \(1120\) \\
North Fork Middle Tule River & \(36.201\) & \(-118.759\) & \(926\) \\
Little Jamison Creek & \(39.743\) & \(-120.704\) & \(1603\) \\
Rock Creek & \(43.374\) & \(-122.957\) & \(326\) \\
\bottomrule
\end{longtable}

@@

\[V_G = 4~\sigma_{\mathrm{parent}}\]
Using \(\sigma_{\mathrm{parent}}\) rather than \(\sigma_{\mathrm{sire}}\) is more powerful because it uses all of the data. We estimated \(\sigma_{\mathrm{parent}}\) and other parameters using a Bayesian mixed effects model. We describe the general approach here and provide detail specific to germination rate and survival below. We fit the model in \emph{Stan} version \(2.29.2\) (\protect\hyperlink{ref-stan_development_team_stan_2022}{Stan Development Team, 2022}) using \textbf{cmdstanr} version \(0.5.2\) (\protect\hyperlink{ref-gabry_cmdstanr_2022}{Gabry and Češnovar, 2022}). \emph{Stan} calculates the posterior distribution of the model using the Hamiltonian Monte Carlo algorithm, which is similar to the more widely used Markov Chain Monte Carlo, but is faster and more efficient at sampling for many applications (\protect\hyperlink{ref-monnahan_faster_2017}{Monnahan et al., 2017}). We used weakly informative normally distributed priors for parameters that affect the trait mean (intercepts and coefficients) and half-Cauchy priors for variance parameters. Weakly informative priors are strongly recommended for complex mixed models (\protect\hyperlink{ref-mcelreath_statistical_2016}{McElreath, 2016}). We ran \(4\) parallel chains for \(4000\) warmup iterations, \(4000\) sampling iterations, and a thinning interval of \(4\). This configuration allowed parameters to converge, which we defined as the convergence diagnostic \(\hat{R} < 1.05\) (\protect\hyperlink{ref-vehtari_rank-normalization_2021}{Vehtari et al., 2021}). We inspected posterior predictive plots to assess model adequacy using the \texttt{pp\_check} function from the \emph{R} package \textbf{bayesplot} version \(1.9.0\) (\protect\hyperlink{ref-gabry_visualization_2019}{Gabry et al., 2019}). We used the posterior median for point estimates and calculated uncertainty with the \(95\)\% quantile intervals from the posterior distribution. \emph{Stan} and \emph{R} code are available on GitHub (\url{https://github.com/cdmuir/mimulus-germination}) and will be archived on Zenodo upon publication. Data will be deposited on Dryad upon publication.

@@

We estimated germination rate under benign greenhouse conditions by measuring time to germination, the number of days to visible emergence from the soil. There were \(48\) randomized blocks split evenly between two time cohorts (North cohort: August \(3 \textendash 4\) and South cohort: August \(22 \textendash 24\), \(2015\)) in the University of British Columbia greenhouse (Vancouver, BC, Canada). The North seed cohort went into the North transplant garden (gardens described below) and the South seed cohort went into the South transplant garden. Seed cohorts were staggered so that seedlings were transplanted at the same ontogenetic stage, as described below. We distinguish between cohorts and gardens because germination for each transplant garden was assayed in the greenhouse prior to transplanting. Emergence is a reliable proxy for germination since seeds were sown directly on top of the soil, resulting in little delay between germination and visible cotyledons. We sowed \(3 \textendash 5\) seeds per plug on moist perlite-peat moss potting medium (Sunshine \#1, Sungro Horticulture, Agawam, Massachusetts, USA) and recorded the first day on which a plug had at least one germinated seedling. We later thinned each plug to one seedling by selecting the central-most individual, so the first germinated seedling recorded is generally not the same individual transplanted into the field. Hence, individual correlations between germination and survival are inappropriate and we analyze relationships between breeding values or population averages. We censused germination daily for most of the experiment, but census dates were more spread out later in the experiment when only a few plants remained. We accounted for census interval in our statistical model (see below). Gentle misting kept soil moist for the duration of the experiment. Most plugs (\(9320\) of \(10650\), \(87.5\)\%) had at least one germinant by the end of the experiment (North cohort: August 27, 2015; South cohort: September 20, 2015); we treated plants that did not germinate as missing data. \(5\) (\(0.05\)\%) plants emerged before day \(5\). We removed data on these individuals because they were likely not \emph{M. cardinalis}, but rather contamination from another species' seed in the potting medium. Species can be difficult to tell apart as the cotyledons first emerge, but in plugs where a single germinant emerged before day \(5\) it was always a contaminant species. It therefore seems most likely that all germinants before day \(5\) were contamination, but observations during thinning showed that contamination was rare and it unlikely to have any substantial impact on our estimate of germination timing.

@@

After germination, we transplanted \(20\) out of \(24\) blocks of seedlings each to two experimental gardens located at the northern and southern portions of the species' range. We refer to these as the ``North'' and ``South'' gardens, respectively. We used some plants from the remaining \(4\) blocks to replace those that did not germinate or died during transport. We transplanted seedlings at similar ontogenetic stages but temporally staggered. We planted the North and South cohorts \(5 \textendash 6\) weeks after germination on September \(9 \textendash 17\), \(2015\) (North seed cohort into North garden) and October \(2 \textendash 9\), \(2015\) (South seed cohort into South garden), respectively. We chose times when we observed natural seedlings were of similar size. The North garden was located near natural populations along the Middle Fork of the Feather River (\(39\)° \(46\)' \(59.7\)" N, \(120\)° \(38\)' \(31.1\)" W, \(1314\) mas, Plumas County, California, USA); the South garden was located along King Creek (\(32\)° \(54\)' \(23.4\)" N, \(116\)° \(37\)' \(26.0\)" W, \(1011\) mas, San Diego County, California, USA). These gardens are located close to two of the source populations (Fig. \ref{fig:range-map}). The North garden is \(7.1\) km from the Little Jamison Creek source population; the South garden is \(3.8\) km from Sweetwater River source population. Since \textit{M. cardinalis} is a riparian specialist, gardens were located within \(50\) m of natural waterways where the microclimate closely resembled that of nearby natural populations. We irrigated plots with microperforated drip tape (Toro, Bloomington, Minnesota) placed about each seedling to help establish transplants and mimic riparian soil where \textit{M. cardinalis} naturally germinates. Seedlings were initially irrigated once per day. The duration of watering was adjusted at each garden to give each plant approximately the same volume of water. Once plants were established, we gradually reduced irrigation as evaporation declined while preventing drought stress. Irrigation ended when winter precipitation began. We discarded data on plants that died within the first month after transplanting, as this probably indicates transplant shock rather than natural mortality. We monitored reemergence in the spring using monthly censuses beginning on March \(14\), \(2016\) (South) and April \(15\), \(2016\) (North). If a plant was recorded in the last census of \(2015\) and emerged in spring \(2016\), we counted it as alive. If a plant did not reemerge, we recorded winter mortality.

@@

Roger Wynn, Gail Wynn, Christina Schneider, and Norbert Schneider provided land for the gardens. Matthew Bayly, Emily Gercke, Kevin Muir, Rachel Muir, Erin Warkman, and Adam Wilkinson helped with data collection or transplanting. Seema Sheth provided seeds and gave feedback on an earlier draft of the manuscript. Lila Fishman and Kelsey Byers provided suggestions on the non-adaptive cline literature.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Selection on early survival does not explain germination rate clines in Mimulus cardinalis},
  pdfauthor={Christopher D. Muir\^{}\{1,2,*\}, Courtney L. Van Den Elzen\^{}\{1,3\}, Amy L. Angert\^{}\{1,4\}},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{geometry}
\geometry{verbose,letterpaper,margin=2.45cm}

@@

\usepackage{setspace}
% \doublespacing

@@

\title{Selection on early survival does not explain germination rate clines in \emph{Mimulus cardinalis}}
\author{Christopher D. Muir\(^{1,2,*}\), Courtney L. Van Den Elzen\(^{1,3}\), Amy L. Angert\(^{1,4}\)}
\date{}

@@

Populations within species differ in ecologically important traits that often result from adaptation to different local environments (\protect\hyperlink{ref-turesson_genotypic_1922}{Turesson, 1922}; \protect\hyperlink{ref-clausen_experimental_1948}{Clausen et al., 1948}). However, it is still rare that we understand the traits and selective agents underlying local adaptation (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}). The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment (\protect\hyperlink{ref-kawecki_conceptual_2004}{Kawecki and Ebert, 2004}; \protect\hyperlink{ref-johnson_reciprocal_2021}{Johnson et al., 2021}), such that populations have higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower \emph{Mimulus guttatus} (Phrymaceae) have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes (\protect\hyperlink{ref-lowry_widespread_2010}{Lowry and Willis, 2010}). The prevalence of local adaptation implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales (e.g. \protect\hyperlink{ref-wright_indirect_2013}{Wright et al., 2013}; \protect\hyperlink{ref-grant_40_2014}{Grant and Grant, 2014}; \protect\hyperlink{ref-richardson_microgeographic_2014}{Richardson et al., 2014}; \protect\hyperlink{ref-lescak_evolution_2015}{Lescak et al., 2015}; \protect\hyperlink{ref-barrett_linking_2019}{Barrett et al., 2019}; \protect\hyperlink{ref-divittorio_natural_2020}{DiVittorio et al., 2020}) likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent (\protect\hyperlink{ref-leimu_meta-analysis_2008}{Leimu and Fischer, 2008}; \protect\hyperlink{ref-hereford_quantitative_2009}{Hereford, 2009}; \protect\hyperlink{ref-brady_understanding_2019}{Brady et al., 2019}) and may be getting weaker because of anthropogenic climate change (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). For example, in the alpine plant \emph{Boechera strica}, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations (\protect\hyperlink{ref-anderson_climate_2020}{Anderson and Wadgymar, 2020}). Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

@@

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}), but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ among environments. If genetic variation in a trait \emph{causes} fitness to vary, then genotypic selection analyses (\protect\hyperlink{ref-rausher_measurement_1992}{Rausher, 1992}) should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of \emph{Arabidopsis thaliana} and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions (\protect\hyperlink{ref-postma_early_2016}{Postma and Ågren, 2016}). Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those that may have evolved nonadaptively.

@@

\emph{Mimulus cardinalis} (\(=\)\emph{Erythranthe cardinalis} \protect\hyperlink{ref-lowry_case_2019}{Lowry et al., 2019}) is a perennial forb native to the Western US (California and Oregon). It is self-compatible, predominantly outcrossing, and hummingbird pollinated. We used five focal populations from throughout the geographic range of \emph{M. cardinalis} (Table \ref{tab:focal_pops}; Appendix S1, Fig. \ref{fig:range-map}; see the Supplementary Data with this article). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits (\protect\hyperlink{ref-muir_grow_2017}{Muir and Angert, 2017}). Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for \(2 \textendash 4\) weeks to dry, then stored at room temperature. To minimize maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.

@@

\[V_G = 4~\sigma_{\mathrm{parent}}\]
Using \(\sigma_{\mathrm{parent}}\) rather than \(\sigma_{\mathrm{sire}}\) is more powerful because it uses all of the data. We estimated \(\sigma_{\mathrm{parent}}\) and other parameters using a Bayesian mixed effects model. We describe the general approach here and provide detail specific to germination rate and survival below. We fit the model in \emph{Stan} version \(2.29.2\) (\protect\hyperlink{ref-stan_development_team_stan_2022}{Stan Development Team, 2022}) using \textbf{cmdstanr} version \(0.5.2\) (\protect\hyperlink{ref-gabry_cmdstanr_2022}{Gabry and Češnovar, 2022}). \emph{Stan} calculates the posterior distribution of the model using the Hamiltonian Monte Carlo algorithm, which is similar to the more widely used Markov Chain Monte Carlo, but is faster and more efficient at sampling for many applications (\protect\hyperlink{ref-monnahan_faster_2017}{Monnahan et al., 2017}). We used weakly informative normally distributed priors for parameters that affect the trait mean (intercepts and coefficients) and half-Cauchy priors for variance parameters. Weakly informative priors are strongly recommended for complex mixed models (\protect\hyperlink{ref-mcelreath_statistical_2016}{McElreath, 2016}). We ran \(4\) parallel chains for \(4000\) warmup iterations, \(4000\) sampling iterations, and a thinning interval of \(4\). This configuration allowed parameters to converge, which we defined as the convergence diagnostic \(\hat{R} < 1.05\) (\protect\hyperlink{ref-vehtari_rank-normalization_2021}{Vehtari et al., 2021}). We inspected posterior predictive plots to assess model adequacy using the \texttt{pp\_check} function from the \emph{R} package \textbf{bayesplot} version \(1.8.1\) (\protect\hyperlink{ref-gabry_visualization_2019}{Gabry et al., 2019}). We used the posterior median for point estimates and calculated uncertainty with the \(95\)\% quantile intervals from the posterior distribution.

@@

We estimated germination rate under benign greenhouse conditions by measuring time to germination, the number of days to visible emergence from the soil. There were \(48\) randomized blocks split evenly between two time cohorts (North cohort: August \(3 \textendash 4\) and South cohort: August \(22 \textendash 24\), \(2015\)) in the University of British Columbia greenhouse (Vancouver, Canada). The North seed cohort went into the North transplant garden (gardens described below) and the South seed cohort went into the South transplant garden. Seed cohorts were staggered so that seedlings were transplanted at the same ontogenetic stage, as described below. We distinguish between cohorts and gardens because germination for each transplant garden was assayed in the greenhouse prior to transplanting. Emergence is a reliable proxy for germination since seeds were sown directly on top of the soil, resulting in little delay between germination and visible cotyledons. We sowed \(3 \textendash 5\) seeds per plug on moist perlite-peat moss potting medium (Sunshine \#1, Sungro Horticulture, Agawam, Massachusetts, USA) and recorded the first day on which a plug had at least one germinated seedling. We later thinned each plug to one seedling by selecting the central-most individual, so the first germinated seedling recorded is generally not the same individual transplanted into the field. Hence, individual correlations between germination and survival are inappropriate and we analyze relationships between breeding values or population averages. We censused germination daily for most of the experiment, but census dates were more spread out later in the experiment when only a few plants remained. We accounted for census interval in our statistical model (see below). Gentle misting kept soil moist for the duration of the experiment. Most plugs (\(9320\) of \(10650\), \(87.5\)\%) had at least one germinant by the end of the experiment (North cohort: August 27, 2015; South cohort: September 20, 2015); we treated plants that did not germinate as missing data. \(5\) (\(0.05\)\%) plants emerged before day \(5\). We removed data on these individuals because they were likely not \emph{M. cardinalis}, but rather contamination from another species' seed in the potting medium. Species can be difficult to tell apart as the cotyledons first emerge, but in plugs where a single germinant emerged before day \(5\) it was always a contaminant species. It therefore seems most likely that all germinants before day \(5\) were contamination, but observations during thinning showed that contamination was rare and it unlikely to have any substantial impact on our estimate of germination timing.

@@

After germination, we transplanted \(20\) out of \(24\) blocks of seedlings each to two experimental gardens located at the northern and southern portions of the species' range. We refer to these as the ``North'' and ``South'' gardens, respectively. We used some plants from the remaining \(4\) blocks to replace those that did not germinate or died during transport. We transplanted seedlings at similar ontogenetic stages but temporally staggered. We planted the North and South cohorts \(5 \textendash 6\) weeks after germination on September \(9 \textendash 17\), \(2015\) (North seed cohort into North garden) and October \(2 \textendash 9\), \(2015\) (South seed cohort into South garden), respectively. We chose times when we observed natural seedlings were of similar size. The North garden was located near natural populations along the Middle Fork of the Feather River (\(39\)° \(46\)' \(59.7\)" N, \(120\)° \(38\)' \(31.1\)" W, \(1314\) mas, Plumas County, California, USA); the South garden was located along King Creek (\(32\)° \(54\)' \(23.4\)" N, \(116\)° \(37\)' \(26.0\)" W, \(1011\) mas, San Diego County, California, USA). These gardens are located close to two of the source populations (Appendix S1, Fig. \ref{fig:range-map}). The North garden is \(7.1\) km from the Little Jamison Creek source population; the South garden is \(3.8\) km from Sweetwater River source population. Since \textit{M. cardinalis} is a riparian specialist, gardens were located within \(50\) m of natural waterways where the microclimate closely resembled that of nearby natural populations. We irrigated plots with microperforated drip tape (Toro, Bloomington, Minnesota, USA) placed about each seedling to help establish transplants and mimic riparian soil where \textit{M. cardinalis} naturally germinates. Seedlings were initially irrigated once per day. The duration of watering was adjusted at each garden to give each plant approximately the same volume of water. Once plants were established, we gradually reduced irrigation as evaporation declined while preventing drought stress. Irrigation ended when winter precipitation began. We discarded data on plants that died within the first month after transplanting, as this probably indicates transplant shock rather than natural mortality. We monitored reemergence in the spring using monthly censuses beginning on March \(14\), \(2016\) (South) and April \(15\), \(2016\) (North). If a plant was recorded in the last census of \(2015\) and emerged in spring \(2016\), we counted it as alive. If a plant did not reemerge, we recorded winter mortality.

@@

Roger Wynn, Gail Wynn, Christina Schneider, and Norbert Schneider provided land for the gardens. Matthew Bayly, Emily Gercke, Kevin Muir, Rachel Muir, Erin Warkman, and Adam Wilkinson helped with data collection or transplanting. Seema Sheth provided seeds and gave feedback on an earlier draft of the manuscript. Lila Fishman and Kelsey Byers provided suggestions on the non-adaptive cline literature. Stephen Weller and two anonymous provided helpful feedback on an earlier version of this manuscript.

@@

\clearpage
